---
title: "Are we more civilized today?"
author: "Nikoline Koldbaek Kaiser"
date: "`r paste('Last updated:', format(Sys.time(), '%d %B %Y, %H:%M'))`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Explore Homicide rates in NW Europe

In this exercise, I load Manuel Eisner's dataset that captures the developments in homicides from the Middle Ages to the 20th century. The goal is to explore the data and create visualisations that show whether homicide rates in Western Europe are lower or higher today than in the past.

This exercise is based on the dataset provided by the [Our World in Data project](https://ourworldindata.org/homicides), based at the University of Oxford.

## The long-term trend in Homicides in Western Europe

Understanding how homicide rates have changed prior to the modern era requires the help of historians and archivists. Manuel Eisner, a criminology professor at the University of Cambridge, and his colleagues published the Historical Violence Database: a compilation of data on long-term trends in homicide rates, in addition to qualitative information such as the cause of death, perpetrator and victim.

Homicide rates are measured as the number of homicides per 100,000 individuals. The data up to 1990 are sourced from Eisner's publication and the Historical Violence Database, while the recent data are taken from the Institute for Health Metrics and Evaluation.

```{r load-libraries}
library(tidyverse)
```

## Load the available data from ourworldindata.org

```{r load-homicide-data}
Western_Europe <- read_csv("data/homicide-rates-across-western-europe.csv")
```

## Inspect the data

The dataset is fairly clean and analysis-ready. `Entity` identifies the country or region, `Year` identifies the year of the observation, and the homicide-rate column gives the number of homicides per 100,000 people. A rate is useful because it adjusts for different population sizes; a raw homicide number would count the total number of deaths without adjusting for how many people lived in the country or region.

```{r inspect-homicide-data}
head(Western_Europe)
names(Western_Europe)
names(Western_Europe)[4] <- "homicides_per_100k"
names(Western_Europe)
glimpse(Western_Europe)
```

## Long-term homicide trends

```{r plot-homicide-trends, eval=TRUE}
Western_Europe %>%
  ggplot(aes(x = Year, y = homicides_per_100k, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  labs(
    title = "Long-term homicide rates in Western Europe",
    x = "Year",
    y = "Homicides per 100,000 people",
    color = "Country"
  ) +
  theme_minimal()
```

The overall pattern is strongly downward. In the medieval and early modern periods, homicide rates were much higher, especially in Italy, Scandinavia, and Germany/Switzerland. By the twentieth and twenty-first centuries, the rates in all five regions are much lower and closer together.

## Individual country trends

```{r facetted-homicide-trends, eval=TRUE}
final_homicide_plot <- Western_Europe %>%
  ggplot(aes(x = Year, y = homicides_per_100k, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.3) +
  facet_wrap(~ Entity, ncol = 2, nrow = 3) +
  labs(
    title = "Homicide rates declined sharply across Western Europe",
    subtitle = "Rates are measured as homicides per 100,000 people",
    x = "Year",
    y = "Homicides per 100,000 people",
    color = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

final_homicide_plot
```

The facetted plot makes the regional differences clearer. Italy starts with the highest homicide rates and also shows the steepest long-term drop. England, the Netherlands, Germany/Switzerland, and Scandinavia all show the same broad direction: a movement from historically high homicide rates to much lower modern rates.

## Compare homicide trends with Danish rulers

```{r load-danish-kings-data, eval=TRUE}
kings <- read_csv2("data/Kongeraekken2.csv")

kings_duration <- kings %>%
  mutate(
    Regerings_start = as.numeric(Regerings_start),
    Regering_slut = as.numeric(Regering_slut),
    duration = Regering_slut - Regerings_start,
    midyear = Regering_slut - ((Regering_slut - Regerings_start) / 2)
  ) %>%
  filter(!is.na(duration), !is.na(midyear), duration >= 0)

head(kings_duration)
```

```{r plot-danish-reign-duration, eval=TRUE}
kings_duration %>%
  ggplot(aes(x = midyear, y = duration)) +
  geom_point(color = "#2f6f73", alpha = 0.75) +
  geom_smooth(color = "#8f2d56", se = FALSE) +
  labs(
    title = "Duration of reign among Danish monarchs",
    x = "Middle year of reign",
    y = "Duration of reign in years"
  ) +
  theme_minimal(base_size = 12)
```

The Danish monarchy plot is more uneven than the homicide plot. Reign duration fluctuates because dynastic politics, wars, deaths, abdications, and succession arrangements all affect individual rulers. Still, compared with the homicide data, the smoother suggests somewhat more stability in later centuries. The homicide plots show a much clearer and stronger decline in violence over time.

## Final tasks

The final visualization I would submit is the facetted homicide plot above, because it directly answers the main question and makes the regional pattern easy to compare.

### Final question

On the basis of these visualisations, Europe does appear to have become more “civilised” if civilisation is defined narrowly as a reduction in lethal interpersonal violence. The homicide data reveal a dramatic long-term decline across England, Italy, Germany and Switzerland, the Netherlands and Belgium, and Scandinavia. During the medieval and early modern periods, homicide rates were extremely high, often reaching levels many times greater than those recorded today. Over the centuries, however, the rates steadily decreased. By the twentieth and twenty-first centuries, the trends in all regions converge at very low levels, usually close to one homicide per 100,000 people or even lower. This suggests that everyday life gradually became less likely to end in deadly violence and that societies developed stronger forms of social control, legal authority, and conflict resolution.

However, the interpretation depends heavily on how the term “civilised” is understood. Homicide statistics measure only one type of violence and cannot account for other forms of suffering, including war, colonial oppression, domestic abuse, inequality, or violence carried out by states. In addition, the Danish reign-duration visualisation shows that political stability has not followed a perfectly consistent path of progress. Therefore, while modern Western Europe may appear more civilised in the specific sense that homicide became far less common, the visualisations do not prove that society became morally superior in every respect. Overall, the evidence points to a major historical decline in recorded lethal violence, but it does not demonstrate the disappearance of cruelty, injustice, or conflict from human society.

